---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Loading libraries

```{r}

library(ggh4x)
library(vegan)
#library(lme4)
library(broom.mixed)
library(tidymodels)
library(ggrepel)
library(dendextend)
library(rstatix)
library(performance)
library(tidyverse)

```


## Functions


```{r}

## Classify tranformation


classify_transformation <- function(formula_name) {
  # Safe element detection regexes for formulas (case-sensitive)
  # Matches the capital letter only if not surrounded by lowercase letters 
  # (avoids matching elements like Na, Si, Se, Pb, etc., or lowercase words)
  is_sulfur_formula     <- "(?<![a-z])S(?![a-z])"
  is_phosphorus_formula <- "(?<![a-z])P(?![a-z])"
  is_nitrogen_formula   <- "(?<![a-z])N(?![a-z])"
  
  case_when(
    # 1. Sulfur-containing (Both formulas with 'S' and sulfur-related names)
    str_detect(formula_name, is_sulfur_formula) |
      str_detect(formula_name, regex("sulfat|thio|cysteine|cystine|glutathione|methionine", 
                                     ignore_case = TRUE)) ~ "Sulfur",
    
    # 2. Phosphorus-containing (Both formulas with 'P' and phosphorus-related names)
    str_detect(formula_name, is_phosphorus_formula) |
      str_detect(formula_name, regex("PO4|phosph|adenylat", ignore_case = TRUE)) ~ "Phosphorus",
    
    # 3. Nitrogen-containing (Both formulas with 'N' and nitrogen-related names)
    str_detect(formula_name, is_nitrogen_formula) |
      str_detect(formula_name, regex("NO3|NH4|urea|amine|amino|uracil|adenine|guanine", 
                                     ignore_case = TRUE)) ~ "Nitrogen",
    
    # 4. Named amino acids
    str_detect(formula_name, regex(
      "alanine|arginine|aspara|aspart|cysteine|glutam|glycine|histidine|isoleucine|leucine|lysine|methionine|phenylalanine|proline|serine|threonine|tryptophan|tyrosine|valine",
      ignore_case = TRUE)) ~ "Amino_acid",
    
    # 5. Carbon-framework (Starts with 'C' in formulas, or C-centric names)
    str_detect(formula_name, "^C(?![a-z])") |
      str_detect(formula_name, regex("sugar|glucose|cellulose|lignin|lipid|fatty", 
                                     ignore_case = TRUE)) ~ "Carbon_framework",
    
    # Fallback
    TRUE ~ "Other"
  )
}


```



# Loading dataset

## Transformations results


```{r}

trans_summary <- read_csv("output/Transformations_summary_all.csv") %>% 
  select(!c("Group", "Transformation", "mf")) %>% 
  mutate(Formula = case_when(
    
    # Matches #N/A-1 through #N/A-5
    str_detect(Formula, "#N/A-[1-5]") ~ str_replace(Formula, "#N/A-", "NA_"),
    TRUE ~ Formula
  ))

trans_counts <- read_csv("output/Transformations_summary_counts.csv") %>% 
  mutate(Formula = case_when(
    
    # Matches #N/A-1 through #N/A-5
    str_detect(Formula, "#N/A-[1-5]") ~ str_replace(Formula, "#N/A-", "NA_"),
    TRUE ~ Formula
  ))

biotic_abio <- read_csv("metabodirect/data/Biotic-abiotic-transfromation-classification.csv")

## Fixing name mismatches

key_fixed <- biotic_abio %>% 
  mutate(
    Name = str_replace_all(Name, "'", "_"),
    
    Name = case_when(
      str_detect(Name, "PO43-")  ~ "PO43",
      str_detect(Name, "SO42-")  ~ "SO42",
      str_detect(Name, "S2O32-") ~ "S2O32",
      str_detect(Name, "NO3-")   ~ "NO3",
      str_detect(Name, "NH4\\+") ~ "NH4",
      
      # Matches #N/A-1 through #N/A-5
      str_detect(Name, "#N/A-[1-5]") ~ str_replace(Name, "#N/A-", "NA_"),
      
      TRUE ~ Name
    )
  )



```

## Create ICR metadata

## Loading


```{r}

icr_metadata <- trans_counts %>%
  filter(!str_detect(SampleID, 'Blk')) %>%
  distinct(SampleID) %>%
  mutate(
    Parent_ID = str_extract(SampleID, "RC2_\\d+"),
    treatment = case_when(
      str_detect(SampleID, "ICR") ~ "Water",
      TRUE ~ "Sediment"),
    SampleID = str_remove_all(SampleID, "SIR-|_p.*"))

#write_csv(icr_metadata, "output/ICR_metadata_6-26-26.csv")


icr_metadata_ft <- icr_metadata 

## Loading tranformations counts raw and normalized by total number of peaks

## Filter transformations counts

trans_counts_ft <- trans_counts 
#%>% 
 # mutate(SampleID = str_remove(SampleID, "SIR-")) %>% 
  #filter(SampleID %in% icr_metadata_ft$SampleID)


```

## Prevalence filtering

```{r}


## Filter singletons and doubletons

ft_mx <- trans_counts_ft %>% 
  filter(Counts > 2) %>% 
  select(SampleID, Formula, Counts) %>% 
  pivot_wider(names_from = "Formula", values_from = "Counts", values_fill = 0) %>% 
  column_to_rownames(var = "SampleID")

## Prevalence filtering

raw_matrix_ft <- ft_mx[,colSums(ft_mx > 5)>=nrow(ft_mx)*0.2]

```



## Tranformations normalization

# Loading CoreMS results

```{r}

files = list.files(path = "CoreMS_input", pattern = ".csv", full.names = TRUE)


```

# Calculate total number of peaks per sample. samples are all in files we need to keep file name as sample ID in the final dataframe

```{r}
# Create an empty dataframe to store the results
total_peak <- data.frame(SampleID = character(length(files)), num_peak = numeric(length(files)), stringsAsFactors = FALSE)

for (i in 1:length(files)){
  sample = read_csv(files[i])
# Extract sampleID from file name for example we need CM_001_ICR-1 from "CoreMS_input/CM_001_ICR-1_p13.corems.csv" or CM_001_SED-1 from "CoreMS_input/CM_001_SED-1_p13.corems.csv"
  total_peak$SampleID[i] = str_extract(files[i], "RC2_\\d+_(ICR|SED)-\\d+|SSS\\d+_(ICR|SED)-\\d+|CM_(ICR|SED)_Blk-\\d+")
  total_peak$num_peak[i] = nrow(sample) 
}


write_csv(total_peak, "output/Total_num_peaks_7-2-26.csv")


```


# Normalize tranformations counts based on total peaks per sample


```{r}

trans_counts_norm <- trans_counts %>%
  mutate(SampleID = str_remove(SampleID, "SIR-|_p.*")) %>%
  left_join(total_peak, by = 'SampleID') %>%
  select(!c(Group, Transformation, Perc_Counts)) %>%
  group_by(SampleID, Formula) %>%
  mutate(norm_count = Counts/num_peak) %>%
  left_join(key_fixed, by = c("Formula" = "Name"))
# 
 write_csv(trans_counts_norm, "output/Transformation_count_normalized_7-2-26.csv")

```



## Normalization by each peak in the tranformation

```{r}


## Adding transformations counts

trans_total <- trans_counts_norm %>%
  group_by(SampleID) %>%
  summarise(
    num_peak = first(num_peak),
    n_transformations = sum(Counts, na.rm = TRUE),
    normalized_transformations = n_transformations / num_peak,
    .groups = "drop"
  )

write_csv(trans_total, "output/Total_Transformations_normalized_7_2_26.csv")



```



## Exploring patterns -- Box plots


```{r}

  
  total_transf <- trans_total %>%
    left_join(icr_metadata_ft, by = "SampleID") %>% 
    filter(!is.na(Parent_ID)) 
  
  pd <- position_dodge2(width = 0.75, preserve = "single")
  
  pj <- position_jitterdodge(
    dodge.width  = 0.75,
    jitter.width = 0.12,
    jitter.height = 0,
    seed = 123
  )
  
  
  box_trans <- total_transf %>%
    mutate(
      Parent_ID = as.factor(Parent_ID),
      treatment = as.factor(treatment)
    ) %>%
    ggplot(aes(x = Parent_ID, y = n_transformations, fill = treatment, colour = treatment)) +
    geom_boxplot(
      width = 0.6,
      alpha = 0.55,
      linewidth = 0.45,
      outlier.shape = NA,
      position = pd
    ) +
    geom_point(
      position = pj,
      size = 1.8,
      alpha = 0.75,
      shape = 16
    ) +
    scale_color_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.08)),
      breaks = scales::breaks_pretty(n = 6)
    ) +
    labs(
      x = "Parent ID",
      y = 'Total number of transformations',
      fill = "Treatment",
      colour = "Treatment"
    ) +
    theme_classic(base_size = 8) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line = element_line(linewidth = 0.45),
      axis.ticks = element_line(linewidth = 0.45),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3),
      panel.grid.major.x = element_blank()
    )
  

box_trans

  
  box_trans2 <- total_transf %>%
    mutate(
      Parent_ID = as.factor(Parent_ID),
      treatment = as.factor(treatment)
    ) %>%
    ggplot(aes(x = Parent_ID, y = normalized_transformations, fill = treatment, colour = treatment)) +
    geom_boxplot(
      width = 0.6,
      alpha = 0.55,
      linewidth = 0.45,
      outlier.shape = NA,
      position = pd
    ) +
    geom_point(
      position = pj,
      size = 1.8,
      alpha = 0.75,
      shape = 16
    ) +
    scale_color_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.08)),
      breaks = scales::breaks_pretty(n = 6)
    ) +
    labs(
      x = "Parent ID",
      y = 'Total number of tranformations normalized by peaks number',
      fill = "Treatment",
      colour = "Treatment"
    ) +
    theme_classic(base_size = 8) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.line = element_line(linewidth = 0.45),
      axis.ticks = element_line(linewidth = 0.45),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3),
      panel.grid.major.x = element_blank()
    )
  
  box_trans2 


## Density plot

  dens_plot <- total_transf %>%
    mutate(treatment = as.factor(treatment)) %>%
    ggplot(aes(x = n_transformations, fill = treatment, colour = treatment)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_rug(aes(colour = treatment), alpha = 0.6, length = unit(0.03, "npc")) +
    scale_color_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_x_continuous(
      breaks = scales::breaks_pretty(n = 6)
    ) +
    labs(
      x = 'Total number of transformations',
      y = "Density",
      fill = "Treatment",
      colour = "Treatment"
    ) +
    theme_classic(base_size = 8) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.line = element_line(linewidth = 0.45),
      axis.ticks = element_line(linewidth = 0.45),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3)
    )

  
dens_plot

  dens_plot2 <- total_transf %>%
    mutate(treatment = as.factor(treatment)) %>%
    ggplot(aes(x = normalized_transformations, fill = treatment, colour = treatment)) +
    geom_density(alpha = 0.4, linewidth = 0.6) +
    geom_rug(aes(colour = treatment), alpha = 0.6, length = unit(0.03, "npc")) +
    scale_color_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_fill_manual(
      values = c(
        "Sediment" = "#D55E00",
        "Water" = "#0072B2"
      )
    ) +
    scale_x_continuous(
      breaks = scales::breaks_pretty(n = 6)
    ) +
    labs(
      x = 'Total number of tranformations normalized by peaks number',
      y = "Density",
      fill = "Treatment",
      colour = "Treatment"
    ) +
    theme_classic(base_size = 8) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.line = element_line(linewidth = 0.45),
      axis.ticks = element_line(linewidth = 0.45),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.3)
    )

  
dens_plot2

```


################################################################################
## Network indices
################################################################################

## Loading netwrok indices


```{r}


nets_stats_df <- read_csv("output/Netwrok_stats_processed_6-1-2026.csv")


net_stats_ready <- nets_stats_df %>% 
  mutate(SampleID = str_remove(SampleID, "SIR-")) %>% 
  left_join(icr_metadata_ft) %>% 
  filter(!is.na(Parent_ID)) %>% 
  split(.$parameters)



indx_plots <- imap(net_stats_ready, function(idx, name){
  
  net_stats_plot <- idx %>% 
    filter(!parameters %in% c("num.neg.edges", "num.pos.edges")) %>% 
    ggplot(aes(x = Parent_ID,
               y = value,
               color = treatment))+
    geom_boxplot()+
    scale_color_manual(
      values = c(
        "DRY" = "#D55E00",
        "WET" = "#0072B2")
    )+
    #facet_wrap(~treatment)+
    geom_point(position = position_jitter())+
    labs(y = name)+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(net_stats_plot)})


indx_plots$num.edges


iwalk(indx_plots, function(plot, name) {
  filename <- paste0("output/figures/", name, "_net_indx.png")
  ggsave(filename, plot, dpi = 300)
})


```


## Netwrok indices grouped by treatment

```{r}


indx_plots2 <- imap(net_stats_ready, function(idx, name){
  
  net_stats_plot <- idx %>% 
    filter(!parameters %in% c("num.neg.edges", "num.pos.edges")) %>% 
    ggplot(aes(x = treatment,
               y = value))+
    geom_boxplot()+
    # scale_color_manual(
    #   values = c(
    #     "DRY" = "#D55E00",
    #     "WET" = "#0072B2")
    # )+
    #facet_wrap(~treatment)+
    geom_point(aes(color = Parent_ID),
               position = position_jitter())+
    labs(y = name)+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(net_stats_plot)})


indx_plots2$centralization.degree



```

## NMDS based on network indices

```{r}


vars_use <- c(
  #"connectance",
  "average.degree",
  "clustering.coefficient",
  "average.path.length",
  #"edge.connectivity",
  #"no.clusters",
  "centralization.degree"
)


nets_stats_wide <- nets_stats_df %>% 
  mutate(SampleID = str_remove(SampleID, "SIR-")) %>% 
  left_join(icr_metadata_ft) %>% 
  filter(!is.na(Parent_ID),
         parameters %in% vars_use) %>% 
  select(SampleID, Parent_ID, treatment, parameters, value) %>% 
  distinct() %>% 
  pivot_wider(names_from = parameters, values_from = value) %>% 
  drop_na(all_of(vars_use))


## Testing for collinearity net indices

dummy_lm <- lm(
  dummy ~ .,
  data = nets_stats_wide %>%
    select(where(is.numeric)) %>%
    mutate(dummy = 1)
)

performance::check_collinearity(dummy_lm) %>%
plot() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))


## Build matrix for ordination


net_meta <- nets_stats_wide %>% 
  select(SampleID, Parent_ID, treatment)


net_matrix <- nets_stats_wide %>% 
  select(SampleID, all_of(vars_use)) %>% 
  column_to_rownames(var = "SampleID") %>% 
  decostand(method = "standardize")


## NMDS (Euclidean distance on standardized network indices)


set.seed(123)
nmds_net <- metaMDS(
  net_matrix,
  distance = "euclidean",
  k = 2,
  trymax = 200,
  autotransform = FALSE,
  trace = FALSE
)


nmds_net_scores <- as_tibble(scores(nmds_net, display = "sites"), rownames = "SampleID") %>% 
  left_join(net_meta, by = "SampleID")


ggplot(nmds_net_scores, aes(x = NMDS1, y = NMDS2, color = treatment)) +
  geom_point(size = 2, alpha = 0.9) +
  stat_ellipse(aes(group = treatment), type = "t", linewidth = 0.8, linetype = 2) +
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    label = paste0("Stress = ", round(nmds_net$stress, 3)),
    hjust = 1.1,
    vjust = -0.8,
    size = 4
  ) +
  scale_color_manual(values = c("DRY" = "#D55E00", "WET" = "#0072B2")) +
  labs(
    title = "NMDS of Network Indices",
    subtitle = "Euclidean distance on standardized indices",
    x = "NMDS1",
    y = "NMDS2",
    color = "Treatment"
  ) +
  coord_equal() +
  theme_classic(base_size = 13)


## PERMANOVA: treatment effect while accounting for Parent_ID (blocked permutations)

perm_treatment_blocked <- adonis2(
  net_matrix ~ treatment,
  data = net_meta,
  method = "euclidean",
  permutations = 9999,
  strata = net_meta$Parent_ID,
  by = "terms"
)

perm_treatment_blocked %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  as_tibble()


## Optional companion model to inspect Parent_ID and treatment terms together

perm_terms <- adonis2(
  net_matrix ~ Parent_ID + treatment,
  data = net_meta,
  method = "euclidean",
  permutations = 9999,
  by = "terms"
)

perm_terms %>% 
  as.data.frame() %>% 
  rownames_to_column("term") %>% 
  as_tibble()


## Beta dispersion

disp <- betadisper(
  dist(net_matrix),
  group = net_meta$treatment
)

anova(disp)
permutest(disp)



```


## Testing ratios


```{r}

net_trans_ratio <- nets_stats_df %>% 
  mutate(SampleID = str_remove(SampleID, "SIR-")) %>% 
  left_join(icr_metadata_ft) %>% 
  filter(!is.na(Parent_ID),
         parameters %in% vars_use) %>% 
  group_by(Parent_ID, treatment, parameters) %>% 
  summarise(median_net = median(value)) 


ggplot(net_trans_ratio,
       aes(treatment, median_net,
           group = Parent_ID)) +
  geom_line(aes(color = Parent_ID),
            alpha = 0.4,
            show.legend = FALSE) +
  geom_point(aes(color = Parent_ID),
             size = 2,
             show.legend = FALSE)+
  facet_wrap(~parameters,
             scales = 'free_y') +
  theme_classic()


```



